Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-08-10 ms/manuscript.Rnw
More edits

\subsection*{Stages 6: Species random effects and trait effects}.

@@

Each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). Wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species capable of achieving larger dbh exhibited lower growth-independent hazards.

@@

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values. B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8. Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e. the stochastic chance of death plus effect of low growth.}
\label{fig4}
\end{figure}

\subsection*{Explanatory power}. 
The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fitted to the data with covariates for growth rate in previous census, species traits -- maximum dbh, light demand and wood density, as well as random effects for both species and census-period. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see  SI Appendix S2 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of species traits -- maximum dbh $\psi$, wood density, $\rho$, and light demand, $\upsilon$ -- and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{0.6}\right)^{\alpha_{1}}\left(\frac{\upsilon}{0.7}\right)^{\alpha_{2}}\left(\frac{\psi}{15}\right)^{\alpha_{3}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ is the effect of wood density $\rho$, $\alpha_{2}$ is the effect of light demand, $\upsilon$, and $\alpha_{3}$ is the effect of maximum  dbh, $\psi$. $\alpha_{0,s}$ captures any other species-level residual error that is otherwise not explained by these three species traits. These random effects were modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around the denominator. In this case we centered wood density on 0.6 g cm$^{-3}$, light demand on 0.7 and maximum dbh on 15 cm. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see  SI Appendix S4 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see  SI Appendix S5-S6 for details). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}. After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior. All parameter estimates from the final model are provided in Table S1\textendash S2 andFigs. S3\textendash S5 in  SI Appendix.

>>>>

\subsection*{Stages 6: Species random effects and trait effects}

@@

Each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). Wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species capable of achieving larger dbh exhibited lower growth-independent hazards. All parameter estimates for the final model can be found in Tables S1\textendash S2.

@@

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values. B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8. Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e. the stochastic chance of death plus effect of low growth. Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
\label{fig4}
\end{figure}

\subsection*{Explanatory power} 
The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fitted to the data with covariates for growth rate in previous census, species traits -- maximum dbh, light demand and wood density, as well as random effects for both species and census-period. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see  SI Appendix S2 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of species traits -- maximum dbh $\psi$, wood density, $\rho$, and light demand, $\upsilon$ -- and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{0.6}\right)^{\alpha_{1}}\left(\frac{\upsilon}{0.7}\right)^{\alpha_{2}}\left(\frac{\psi}{15}\right)^{\alpha_{3}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ is the effect of wood density $\rho$, $\alpha_{2}$ is the effect of light demand, $\upsilon$, and $\alpha_{3}$ is the effect of maximum  dbh, $\psi$. $\alpha_{0,s}$ captures any other species-level residual error that is otherwise not explained by these three species traits. These random effects were modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around the denominator. In this case we centered wood density on 0.6 g cm$^{-3}$, light demand on 0.7 and maximum dbh on 15 cm. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see  SI Appendix S4 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see  SI Appendix S5-S6 for details). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}. After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior. All parameter estimates from the final model are provided in Table S1\textendash S2 and Figs. S3\textendash S5 in SI Appendix.

### 2018-08-10 ms/supplementary.Rnw
More edits

Here we provide some further details on methods and additional results, supplementing those presented in the main text. 
The section ``Review of methods for estimating mortality'' 
``Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community. 
The section ``Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. 
The section ``Numerical optimisation'' outlines some numerical concerns in estimating logloss, while 
the section ``Example Stan code'' gives an example of the code fitted to the data. Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.

@@

First, is whether the time interval between census intervals is explicitly included in the model. Some methods model binary survival outcomes directly, while others seek to estimate the underlying mortality rate experienced by a population per unit time. Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011,}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}. These methods are suitable for detecting significant correlations with predictors, or predicting survival over a fixed time interval. However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Kleinbaum:2012tn}. Models formulated around a hazard rate can therefore make predictions across different time intervals. Moreover, the hazard rate connects directly with the differential equations used to build process-based models of vegetation \cite[e.g.][]{Moorcroft:2001ws}. While a number of studies have directly estimated hazard rates, studies differ in the structure of the estimated hazard rate. Many studies estimate average hazard rates for populations of each species \cite[e.g.][]{Condit:2006, king:2006va, Kraft:2010kq, Wright:2010fl}. These studies are effective at quantifying differences between species but unable to explain variation among individuals within species. Others, seek to model variation in the hazard rate via a combination covariates \cite[e.g.][]{Ruger:2011cv} (see also main text).

@@

Studies using logistic regression, or Generalized linear models with a logit link , all seek to predict he probability of death over a specified time interval via a model of the form:
\begin{equation} \label{eq:logistic_model}
p_{i,t_1\rightarrow t_2} = \frac {1}{1+e^{ (\beta _{0} + \beta _{1}x_1 + \ldots )}},
\end{equation}
where $\beta_{0}, \beta _{1}, \dots$ are coefficients and $x_{1}, x_{2}, \dots$ are covariates. As increases in growth tend to decrease mortality risk, the coefficient for growth rate when used as a predictor will be positive. High growth rates will cause the exponential term to become large, and $p_{i,t_1\rightarrow t_2} \rightarrow 0$.

@@

\clearpage
\begin{figure*}[!hb]
\centering
\includegraphics{figures/supp_fig6}
\caption{Pairwise comparison between species' wood density, light demand and maximum dbh.}
\label{figS6}
\end{figure*}
\end{document}

>>>>

Here we provide some further details on methods and additional results, supplementing those presented in the main text. 
The section ``Review of methods for estimating mortality'' briefly describes the statistical approaches commonly used when estimating tree mortality and how they differ to the additive hazards model advocated in this manuscript.
``Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community. 
The section ``Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. 
The section ``Numerical optimisation'' outlines some numerical concerns in estimating logloss, while 
the section ``Example Stan code'' gives an example of the code fitted to the data. Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.

@@

First, is whether the time interval between census intervals is explicitly included in the model. Some methods model binary survival outcomes directly, while others seek to estimate the underlying mortality rate experienced by a population per unit time. Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}. These methods are suitable for detecting significant correlations with predictors, or predicting survival over a fixed time interval. However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Zens:2003jv,Kleinbaum:2012tn}. Models formulated around a hazard rate can therefore make predictions across different time intervals. Moreover, the hazard rate connects directly with the differential equations used to build process-based models of vegetation \cite[e.g.][]{Moorcroft:2001ws}. While a number of studies have directly estimated hazard rates, studies differ in the structure of the estimated hazard rate. Many studies estimate average hazard rates for populations of each species \cite[e.g.][]{Condit:2006, king:2006va, Kraft:2010kq, Wright:2010fl}. These studies are effective at quantifying differences between species but unable to explain variation among individuals within species. Others, seek to model variation in the hazard rate via a combination covariates \cite[e.g.][]{Ruger:2011cv} (see also main text).

@@

Studies using logistic regression, or Generalized linear models with a logit link, all seek to predict the probability of death over a specified time interval via a model of the form:
\begin{equation} \label{eq:logistic_model}
p_{i,t_1\rightarrow t_2} = \frac {1}{1+e^{ (\beta _{0} + \beta _{1}x_1 + \ldots )}},
\end{equation}
where $\beta_{0}, \beta _{1}, \dots$ are coefficients and $x_{1}, x_{2}, \dots$ are covariates. As increases in growth tend to decrease mortality risk, the coefficient for growth rate when used as a predictor will be positive. High growth rates will cause the exponential term to become large, and $p_{i,t_1\rightarrow t_2} \rightarrow 0$.

@@

\clearpage
\begin{figure*}[!hb]
\centering
\includegraphics{figures/supp_fig6}
\caption{Pairwise comparison between species' wood density, light demand and maximum dbh. Correlations are pearson correlations}
\label{figS6}
\end{figure*}
\end{document}
